Bayesian Sparsification of Deep $\cplx$-valued networks

### 2020-02-05 paper/paper.tex
edits and improvements in training proc description

\subsection{Fitting and compression} % (fold)
\label{sub:fitting_and_compression}

The goal of this section to study the compression-performance trade-off of the proposed Bayesian
sparsification methods for $\cplx$-valued networks. To this end we employ the following training
plan in every experiment. The model is trained in three successive stages: ``dense'' $\to$
``sparsify'' $\to$ ``fine-tune''. Parameter values and the internal state of the optimizers
are transferred between stages.

@@

At the \textit{``dense''} stage every network is fit ``as-is'', retaining its original
non-Bayesian architecture and using only the likelihood term from \eqref{eq:elbo} without
variational dropout.
%
During the \textit{``sparsify''} stage we make every layer Bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:variational_dropout} (or sec.~\ref{ssub:empirical_bayes}) or its $\real$
counterpart. To be able to study the attainable sparsity levels we inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo}:
\begin{equation}  \label{eq:elbo_with_coef}
    - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
    + \frac1{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,. \tag{\ref{eq:elbo}'}
\end{equation}
Higher $C$ yields increasingly more sparse layers upon termination of this stage.
%
The \textit{``fine-tune''} stage begins with computing the sparsity masks from the fitted
approximation $q_\theta(W)$ and proceeds to train the non-zero parameters similarly to
the ``dense'' stage, i.e. with non-Bayesian layers and fixed sparsity masks. The test
performance is reported for the model at the start and termination of the final stage.

Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:dropout}). Since
the variational approximation factorizes into univariate distributions, whether a $\cplx$
or $\real$ parameter is zero or not is determined from individual relevance scores $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$: non-zero if $\log \alpha \leq \tau$, zero otherwise. Figure~\ref{fig:log-alpha__mnist-like}
displays the general shape of the distribution of $\log \alpha$ on the Fashion-MNIST dataset
in sec.~\ref{sub:mnist_like_datasets}. This shape appears to be typical in every experiment
conducted in this study: there always is a single mode within the moderately negative
range, and one or two modes to the right of zero. The relative mass of each mode changes
with $C$ and the higher the coefficient is, the more mass is shifted to the right. In every
experiment the trough between the peaks of the negative mode and the first positive one
is around zero.

We decide if a parameter is \textit{``non-zero''} if it is within $\delta$ relative
tolerance of its mean with high probability under the approximate posterior. Since
\eqref{eq:c-gauss-vi} is Gaussian, $
  \tfrac{\lvert w - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with shape $k=1$ ($\real$) or $2$ ($\cplx$), whence
\begin{equation} \label{eq:prob_relevant}
  \Pr\bigl(
    \lvert w - \mu \rvert \leq \delta \lvert \mu \rvert
  \bigr)
    = \Pr\biggl(
      \frac{\lvert w - \mu \rvert^2}{\alpha \lvert \mu \rvert^2}
        \leq \frac{\delta^2}{\alpha}
    \biggr)
    = \chi^2_k\biggl(\frac{\delta^2}{\alpha}\biggr)
  \,.
\end{equation}
% Figure~\ref{fig:log-alpha__mnist-like} (bottom) shows the probability of a $\real$
% (or $\cplx$)  parameter being non-zero for various $\delta\in(0, \tfrac12)$ as a function
% of $\log \alpha$.
For a reasonably generous $\delta = 25\%$ tolerance all values of $\log \alpha$ below
$-2.5$ yield a relatively high chance of a non-zero parameter. Given that $q_\theta$
is merely an approximation, we opt for a somewhat higher $\log \alpha$ threshold, and
pick $\tau = -\tfrac12$, which is much lower than $\tau = 3$, recommended by
\citet{molchanov_variational_2017}. Additional rationale behind this decision is the
extra fine tuning, which trains the compressed model.

@@

We conduct a moderately sized numerical experiment on MNIST and similar $28\times 28$
greyscale image datasets to study the compression-performance trade-off of the proposed
$\cplx$-variational dropout on simple architectures. We use classification accuracy as
the target performance metric. We compare with the $\real$-valued networks against their
$\cplx$-valued counterparts with approximately half as many parameters, as suggested
by \citet{monning_evaluation_2018}.

% dataset size
We deliberately use a {fixed} random subset of $10k$ images from the train split of each
dataset to fit the networks, and gauge the final performance on the usual test split.
Limiting the size was done primarily to test the strength of regularization properties
of variational dropout, and to make experiments quicker.
% image features
Since image data is not naturally $\cplx$-valued, we reprocess images by either embedding
$\real$-data into $\cplx$ assuming $\Re z = 0$ (\texttt{raw}), or applying the two-dimensional
Fourier Transform (\texttt{fft}), shifting lower frequencies to the centre of the image. We
do not investigate the option, proposed by \citet{trabelsi_deep_2017}, to train an auxiliary
model to synthesize the imaginary component from the real data.

% model architectures
We study the following $\real$/$\cplx$ networks:
\begin{itemize}
  \item \textit{SimpleConvModel}: $2d$ convnet with Average pooling ($2\times 2$)
  with layers $c(n_1, 5, 1) \to c(n_2, 5, 1) \to d(n_3) \to d(n_\mathtt{out})$
  \item \textit{TwoLayerDenseModel}: wide shallow network with structure
  $d(n_1) \to d(n_\mathtt{out})$
\end{itemize}
where $c(n, k, s)$ denotes a convolution with $n$ output channels, kernel width $k$ and
stride $s$, and $d(n)$ is a dense layer with $n$ outputs. The values of $n_\cdot$ depend
on the input features and are summarized in table~\ref{tbl:mnist-like-model-spec}.
\begin{table}[hb]
  \centering
  \caption{Architecture parameters for models studied on MNIST-like datasets.}
  \label{tbl:mnist-like-model-spec}
  \begin{tabular}{lrr}
    \toprule
    $(n_1, n_2, n_3)$  & $\real$-\textit{SimpleConvModel} & $\cplx$-\textit{SimpleConvModel} \\
    \midrule
    (\texttt{raw}) & $20, 50, 500$ & $10, 25, 250$ \\
    (\texttt{fft}) & $40, 100, 1000$ & $20, 50, 500$ \\
    \bottomrule
  \end{tabular}
  \begin{tabular}{lrr}
    \toprule
    $(n_1)$  & $\real$-\textit{TwoLayerDenseModel} & $\cplx$-\textit{TwoLayerDenseModel} \\
    \midrule
    (\texttt{raw}) & $4096$ & $2048$ \\
    (\texttt{fft}) & $8192$ & $4096$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

% discussion of TwoLayerDenseModel
% discussion of SimpleConvModel
% hinting at overfitting/regularization effects of masking and sparsification

>>>>

\subsection{Staged fitting and compression} % (fold)
\label{sub:fitting_and_compression}

The goal of this section to study the compression-performance trade-off of the proposed Bayesian
sparsification methods for $\cplx$-valued networks. To this end we employ the following training
plan in every experiment. The model is trained in three successive stages: ``dense'' $\to$
``sparsify'' $\to$ ``fine-tune''. Model parameters and the optimizer are copied between
stages, except before the ``fine-tune'' optimizer's internal state is reset.

At the \textit{``dense''} stage every network is fit ``as-is'', retaining its original
non-Bayesian architecture and using only the likelihood term from \eqref{eq:elbo} without
variational dropout.
%
During the \textit{``sparsify''} stage we make every layer Bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:variational_dropout} (or sec.~\ref{ssub:empirical_bayes}) or its $\real$
counterpart. To be able to study the attainable sparsity levels we inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo}:
\begin{equation}  \label{eq:elbo_with_coef}
    - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
    + \frac1{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,. \tag{\ref{eq:elbo}'}
\end{equation}
Higher $C$ yields increasingly more sparse layers upon termination of this stage.
%
The \textit{``fine-tune''} stage begins with computing the sparsity masks from the fitted
approximation $q_\theta(W)$ and proceeds to train the non-zero parameters similarly to
the ``dense'' stage, i.e. with non-Bayesian layers and fixed sparsity masks.

% on compresssion
Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:dropout}). Since
the variational approximation factorizes into univariate distributions, whether a $\cplx$
or $\real$ parameter is zero or not is determined from individual relevance scores $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$: non-zero if $\log \alpha \leq \tau$, zero otherwise. Figure~\ref{fig:log-alpha__mnist-like}
displays the general shape of the distribution of $\log \alpha$ on the Fashion-MNIST dataset
in sec.~\ref{sub:mnist_like_datasets}. This shape appears to be typical in every experiment
conducted in this study: there always is a single mode within the moderately negative
range, and one or two modes to the right of zero. The relative mass of each mode changes
with $C$ and the higher the coefficient is, the more mass is shifted to the right. In every
experiment the trough between the peaks of the negative mode and the first positive one
is around zero.

% on compression threshold
The threshold $\tau$ is picked so, that the remaining \textit{``non-zero''} parameters
are within $\delta$ relative tolerance of their mean with high probability under the
approximate posterior $q_\theta(W)$. Since \eqref{eq:c-gauss-vi} is Gaussian, $
  \tfrac{\lvert w - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with shape $k=1$ ($\real$) or $2$ ($\cplx$), whence
\begin{equation} \label{eq:prob_relevant}
  \Pr\bigl(
    \lvert w - \mu \rvert \leq \delta \lvert \mu \rvert
  \bigr)
    = \Pr\biggl(
      \frac{\lvert w - \mu \rvert^2}{\alpha \lvert \mu \rvert^2}
        \leq \frac{\delta^2}{\alpha}
    \biggr)
    = \chi^2_k\biggl(\frac{\delta^2}{\alpha}\biggr)
  \,.
\end{equation}
% Figure~\ref{fig:log-alpha__mnist-like} (bottom) shows the probability of a $\real$
% (or $\cplx$)  parameter being non-zero for various $\delta\in(0, \tfrac12)$ as a function
% of $\log \alpha$.
For a reasonably generous $\delta = 25\%$ tolerance all values of $\log \alpha$ below
$-2.5$ yield a relatively high chance of a non-zero parameter. Given that $q_\theta$
is merely an approximation, we opt for a somewhat higher $\log \alpha$ threshold, and
pick $\tau = -\tfrac12$, which is much lower than $\tau = 3$, recommended by
\citet{molchanov_variational_2017}.

\subsection{Metrics and performance} % (fold)
\label{sub:metrics_and_performance}

% perfromance metrics
We use accuracy as the performance metric for multi-class image classification on MNIST-like
and CIFAR10 datasets. Similarly to \citet{trabelsi_deep_2017} the performance in MusicNet
experiments is measured with (pooled) average precision: predicted scores and target binary
labels for each output are pooled, and average precision is computed based on the precision-recall
curve as if the task was single-output.

The test performance is reported befire the beginning and after termination of the ``fine-tune''
stage. Additional rationale behind this decision is the extra fine tuning, which trains the
compressed model.

@@

% subsection metrics_and_performance (end)

@@

We conduct a moderately sized numerical experiment on MNIST and similar $28\times 28$
greyscale image datasets to study the compression-performance trade-off of the proposed
$\cplx$-variational dropout on simple architectures. We compare with the $\real$-valued
networks against their $\cplx$-valued counterparts with approximately half as many
parameters, as suggested by \citet{monning_evaluation_2018}.

% dataset size
We deliberately use a {fixed} random subset of $10k$ images from the train split of each
dataset to fit the networks, and gauge the final performance on the usual test split.
Limiting the size was done primarily to test the regularization strength of sparsification
performed by variational dropout, and to make bulk experimentation faster.

% image features
Since image data is not naturally $\cplx$-valued, we preprocess images by either embedding
$\real$-data into $\cplx$ assuming $\Re z = 0$ (\texttt{raw}), or applying the two-dimensional
Fourier Transform (\texttt{fft}), shifting lower frequencies to the centre of the image. We
do not investigate the option, proposed by \citet{trabelsi_deep_2017}, to train an auxiliary
model to synthesize the imaginary component from the real data.

% model architectures
We study the following $\real$/$\cplx$ networks:
\begin{itemize}
  \item \textit{SimpleConvModel} (SCM): $2d$ convnet with Average pooling ($2\times 2$)
  with layers $c(n_1, 5, 1) \to c(n_2, 5, 1) \to d(n_3) \to d(n_\mathtt{out})$
  \item \textit{TwoLayerDenseModel} (TLDM): wide shallow network with structure
  $d(n_1) \to d(n_\mathtt{out})$
\end{itemize}
where $c(n, k, s)$ denotes a convolution with $n$ output channels, kernel width $k$ and
stride $s$, and $d(n)$ is a dense layer with $n$ outputs. The values of $n_\cdot$ depend
on the input features and are summarized in table~\ref{tbl:mnist-like-model-spec}.
\begin{table}[hb]
  \centering
  \caption{Architecture parameters for models studied on MNIST-like datasets.}
  \label{tbl:mnist-like-model-spec}
  \begin{tabular}{lrr}
    \toprule
    $(n_1, n_2, n_3)$  & $\real$-\textit{SCM} & $\cplx$-\textit{SCM} \\
    \midrule
    (\texttt{raw}) & $20, 50, 500$ & $10, 25, 250$ \\
    (\texttt{fft}) & $40, 100, 1000$ & $20, 50, 500$ \\
    \bottomrule
  \end{tabular}
  \begin{tabular}{lrr}
    \toprule
    $(n_1)$  & $\real$-\textit{TLDM} & $\cplx$-\textit{TLDM} \\
    \midrule
    (\texttt{raw}) & $4096$ & $2048$ \\
    (\texttt{fft}) & $8192$ & $4096$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

% discussion of TwoLayerDenseModel (TLDM) and SimpleConvModel (SCM)
% hinting at overfitting/regularization effects of masking and sparsification
